Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-12-07 manuscript.Rmd
Danny's changes

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      First line
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in. For example, the risk of dying is not independent of time. Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10. This risk can be increased or decreased by behavioral (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural), but there is an underlying risk that is time dependent and impacts on all units.

However, sometimes we are interested in situations in which not all units are at risk of dying, failing, or acquiring some new characteristic. Imagine we are interested in the risk of acquiring the flu. First, we might think that everyone is at risk of getting the flu and it just depends on individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) whether individuals will acquire the flu. If this would be the case we could use a standard duration model and estimate how different factors impact on the baseline risk. But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or some other factors that makes it impossible for them to get sick. Given that this is the case we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.

@@

Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-paramaetric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.

@@

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

@@

The cumulative failure rate ($F(t) = 1 - S(t)$) over time converges to 1, meaning all subjects fail eventually. This assumption is untenable in many situations. Some cancer patients are cured after treatment while others relapse, most young people do not start smoking even though most smokers started at a young age, and many states will not experience the kind of violence that persists in some parts of the world. A better assumption in such contexts is that there is a subpopulation of subjects who are at risk of experiencing an event at some point, and another subpopulation who will never experience the event.

The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured. A model will thus over predict hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.

@@

Crucially, this split-population framework is primarily useful in contexts where sub-populations are not clearly or easily identifiable. Even though there is a clear sub-populations in a model of the age at first pregnancy for humans---men---no researcher in their right mind would think to include male subjects in their data at all, for example. On the other hand, whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark is much less clear and in fact the question that initially led to the idea for split-population duration modeling [@boag1949maximum, @berkson1952survival].

@@

Where $\lambda = e^{-x_i\beta}$ is a parameter of covariates.

@@

For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail. In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, give survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.

@@

The \pkg{spduration} package for \proglang{R} implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, Weibull and log-logistic. The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.

As an example to illustrate the package functionality, we will replicate and extend the model of coup onsets in @belkin2003toward.

Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in a regime at risk. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they say "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [@belkin2003toward, p. 598].

@@

Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced a failure event, until the next failure event or end of observation (right censoring). Table \ref{tab-ex} shows how a single country could have two spells that consist of the observations for that country from 1960, when our data start, to a failure event in say 1980, when a new spell starts and continues until the next failure event or data end. Country "B" on the other hand experienced no coups and all observation for that country from 1960 to 2000 make up one spell.

@@

Table \ref{loglog_table} shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure. The negative coefficient for military regimes for example shows that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant. In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.

@@

Many outcomes are rare events. Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable. Figure \ref{rates} shows a few examples for published research that models binary outcomes. @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries. The positive rates range from 5 to less than 0.2\% of all data points.

>>>>

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill, Jr.
    affiliation: University of Georgia
    address: >
      First line
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      First line
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in. For example, the risk of dying is not independent of time. Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10. Individual behaviors (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural) can increase or decrease this risk, but there is an underlying risk that is time dependent and impacts on all units.

However, sometimes we are interested in situations in which not all units are at risk of dying or experiencing some other event of interest. Imagine we are interested in the risk of acquiring the flu. We might think that everyone is at risk of getting the flu, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. If this were the case we could use a standard duration model and estimate how different factors impact on the baseline risk. But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or have some other characteristic that makes it impossible for them to get sick. Given that this is the case we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.

@@

Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.

@@

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal relapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

@@

The cumulative failure rate ($F(t) = 1 - S(t)$) over time converges to 1, meaning all subjects fail eventually. In the examples of applied research discussed above this assumption is untenable. Some cancer patients are cured after treatment, most young people never become regular smokers, and many contemporary states will not experience the kind of violence that persists in some parts of the world. The presence of a large sub-population which is not at risk for an event in practice will inflate estimates of the survival fraction, and reduce hazard estimates for all subjects by essentially, and falsely, distributing risk among subjects that genuinely will fail and those that are cured. A model will thus over predict hazard for subjects that are not at risk (cured), and under predict for those who are at risk of eventually experiencing the event of interest.

@@

Crucially, this split-population framework is primarily useful in contexts where sub-populations are not clearly or easily identifiable. For example, there is a clear sub-population in a model of the age at first pregnancy for humans---men---which researchers can easily identify and exclude from their data. Much less clear are answers to questions such as whether a cancer patient is cured or not cured given that they have no visible signs of cancer following treatment or have hit the 5-year disease free survival mark. In such situations split-population models provide a way to infer membership in sub-populations in a probabilistic fashion.

@@

Where $\lambda = e^{-x_i\beta}$ is a parameter of covariates. The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.

@@

For a given unconditional risk rate $\pi$, the probability that a particular case with survival time $t$ is in the risk set decreases over time because an increasing number of surviving cases consist of immune or cured ($1-\pi$) cases that will never fail. In the hazard rate, the failure rate in the numerator is conditional on the probability that a case is in the risk set, given survival up to time $t$, and the numerator is an adjusted survivor function that accounts for the fraction of cured cases by time $t$, which is $1-\pi(t)$.

@@

As an example to illustrate the package functionality, we will examine the model of coups d`etat in @belkin2003toward. Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in an at-risk regime. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they say "[t]riggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [@belkin2003toward, p. 598].

@@

Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced the event of interest, until the next event or the end of observation (right censoring). Table \ref{tab-ex} shows how a single country could have two spells during the observation period. Observations for Portugal make up two spells, one that begins in 1960, when our data start, to a coup event in 1975, when a second spell starts and continues until either the next coup or the data end. Canada on the other hand experienced no coups and so observations for that country from 1960 to 2000 make up one spell.

@@

Table \ref{loglog_table} shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure. The negative coefficient for military regimes for example means that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant. In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.

@@

Many outcomes of interest are rare events. Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable. Figure \ref{rates} shows a few examples for published research that models binary outcomes. @fearon2003ethnicity is a widely cited study of civil war onset that uses yearly observations of all major countries; @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries. The positive rates range from 5 to less than 0.2\% of all data points.
